Local Spectral Clustering of Density Upper Level Sets

### 2021-12-23 neurips/local_spectral_density_clustering_arxiv.tex

\begin{theorem}
  \label{thm: consistent_recovery_of_density_clusters}
  Fix $\lambda > 0$, let $\Cset \in \Cbb_f(\lambda)$ be a
  $\kappa$-well-conditioned density cluster (with respect to some $\theta$), and
  additionally assume $f$ satisfies \ref{asmp: C'_bounded_density}. If Algorithm
  \ref{alg: ppr} is well-initialized, there exists a universal constant $c >  0$
  such that if  
  \begin{equation}
    \label{eqn: kappa_ub}
    \kappa(\Cset) \leq c \frac{\lambda_{\sigma}^2r^d
      \nu_d}{\Lambda_{\sigma}\Pbb(\Csig)}, 
  \end{equation}
  then the output set $\Cest \subseteq \Xbf$ is a consistent estimator for
  $\Cset$, in the sense of Definition \ref{def: consistent_density_cluster_estimation}. 
\end{theorem}
\noindent Some remarks are in order: 
 \begin{enumerate}  
 \item We note that the restriction on $\kappa(\Cset)$ imposed by \eqref{eqn:
    kappa_ub} results in a symmetric set difference \smash{$\Delta(\Csig[\Xbf],
    \Cest)$} on the order of $r^d$. In plain terms, we are able to recover a
  density cluster $\Cset$ in the sense of \eqref{eqn:
    consistent_density_cluster_recovery} only when we can guarantee a very small
  fraction of points will be misclassified. This strong condition is the price
  we pay in order to obtain the uniform bound in~\eqref{eqn: ppr_gap}. 
 \item While taking the radius of the neighborhood graph $r \to 0$ as $n \to
  \infty$ (thereby ensuring that the graph $G_{n,r}$ is sparse) is computationally
  attractive, the presence of a factor of \smash{$\log^2(1/r)/r$} in
  $\kappa(\Cset)$ prevents us from making claims about the
  behavior of PPR in this regime. Although the restriction to a kernel
  function fixed in $n$ is standard for spectral clustering theory
  \citep{schiebinger2015,vonluxburg2008}, it is an interesting question whether
  PPR exhibits some degeneracy over $r$-neighborhood graphs as $r \to 0$, or if
  this is merely looseness in our upper bounds.  
\end{enumerate}
\sbcomment{You are using $\epsilon$ here and in the future to mean different things. I have
changed this $\epsilon$ to $\varepsilon$.}

@@

\subsection{Upper Bounds on the Mixing Time} 
For $S \subseteq V$, denote by $G[S] = (S, E_S)$ the
subgraph induced by $S$ (where the edges are $E_S = E \cap (S \times S)$). Let
$\Wbf_S$ be the (lazy) random walk matrix over $G[S]$, and write  
$$
q_{v}^{(t)}(u) = e_v\Wbf_S^t e_u
$$
for the $t$-step transition probability of the lazy random walk over $G[S]$
originating at $v \in V$. Also write \smash{$\pi = (\pi(u))_{u \in S}$}
for the stationary distribution of this random walk.  (As
$\Wbf_S$ is the transition matrix of a lazy random walk, 
it is well-known that a unique stationary distribution exists and is given by 
\smash{$\pi(u) = (\Dbf_S)_{uu}/\vol(S; G[S])$}, where we write $\Dbf_S$ for the  
degree matrix of $G[S]$.) We define the \emph{mixing time} of $G[S]$ as
\begin{equation}
\label{eqn: mixing_time}
\tau_{\infty}(G[S]) = \min\set{ t: \frac{\pi(u) - q_{v}^{(t)}(u)}
  {\pi(u)} \leq \frac{1}{4}, \; \text{for $u,v \in V$}}. 
\end{equation}
Next, we give an asymptotic (in the number of vertices $n$) upper bound on
$\tau_{\infty}(G_{n,r}[\Csig[\Xbf]])$.

\begin{theorem}
  \label{thm: mixing_time_upper_bound}
  Fix $\lambda > 0$, and assume that \smash{$\Cset \in \Cbb_f(\lambda)$}
  satisfies Assumptions \ref{asmp: bounded_density} and \ref{asmp: embedding}
  for some $\sigma, \lambda_{\sigma}, \Lambda_{\sigma}, \rho, L > 0$. Then, for
  any \smash{$0 < r < \sigma/2\sqrt{d}$}, with probability 1,
  \begin{equation} 
    \label{eqn: mixing_time_upper_bound}
    \limsup_{n \to \infty}\tau_{\infty}(G_{n,r}[\Csig[\Xbf]]) \leq c_2
    \frac{\Lambda_{\sigma}^4 d^3 \rho^2 L^2}{\lambda_{\sigma}^4 r^2}
    \log^2\left(\frac{1}{r}\right) + c_3,
  \end{equation}
  for $c_2,c_3 > 0$ universal constants. 
\end{theorem}

@@

\paragraph{Empirical behavior of PPR.} In Figure \ref{fig:fig2}, to drive home
the main implications of Theorems \ref{thm: misclassification_rate} and
\ref{thm: consistent_recovery_of_density_clusters}, we show the
behavior of PPR, normalized cut, and the density clustering algorithm of
\citet{chaudhuri2010} on the well known ``two moons'' dataset (with added 2d 
Gaussian noise), considered a prototypical success story for spectral clustering
algorithms. The first column shows the empirical density clusters $\Cset[\Xbf]$
and $\Cset'[\Xbf]$ for a particular threshold $\lambda$ of the density function; the 
second column shows the cluster recovered by PPR; the third column shows the
global minimum normalized cut, computed according to the algorithm of
\citet{szlam2010}; and the last column shows a cut of the density cluster tree
estimator of \citet{chaudhuri2010}.  We can see the degrading ability of PPR to
recover density clusters as the two moons become less well-separated. Of 
particular interest is the fact that PPR fails to recover one of the moons even
when normalized cut still succeeds in doing so, supporting our claim from Remark
\ref{rmk: diameter}. Additionally, we note that the Chaudhuri-Dasgupta algorithm 
succeeds even when both PPR and normalized cut fail.  While our main message was
that PPR recovers geometrically well-conditioned density clusters, it would be  
interesting to establish that it \emph{only} recovers such clusters, a direction
for future work.

@@

Taking these two technical lemmas as given we can now complete the proof of the theorem. 
We fix $\delta \in (0,1]$ and let $\delta' = \delta/3$. We rewrite
\begin{equation}
\label{eqn: conductance_representation_1}
\Phi_{n,r}(\Csig[\Xbf]) = \frac{p_K + \left(\frac{\cut_{n,r}}{{n \choose 2}} - p_K\right)}{p_V + \left(\frac{\min\set{\vol_{n,r}, \vol_{n,r}^c}}{{n \choose 2}} - p_V\right)}.
\end{equation}
By a union bound we see that all three bounds in \eqref{eqn: numerator_additive_bound} hold with probability at least $1 - \delta$. Along with (\ref{eqn: conductance_representation_1}) this yields the upper bound \sbcomment{Changed $\mathbf{X}$ to $X$}
\begin{equation*}
\Phi_{n,r}(\Csig[X]) \leq \frac{p_K + \Err_n}{p_V - \Err_n}
\end{equation*}
for $\Err_n = \sqrt{\frac{\log(1/\delta')}{n}}$.
Now, some straightforward algebraic manipulations yield,
\begin{align*}
\frac{p_K + \Err_n}{p_V - \Err_n} & = \frac{p_K}{p_V} \left(\frac{p_V}{p_V - \Err_n}\right) + \frac{\Err_n}{p_V - \Err_n} \\
& = \frac{p_K}{p_V} + \left(\frac{p_K}{p_V} + 1\right)\frac{\Err_n}{p_V - \Err_n} \\
& \leq \frac{p_K}{p_V} + 2 \frac{\Err_n}{p_V - \Err_n}.
\end{align*}

@@

\subsection{Proof of Lemma~\ref{lem: expected_density_cut}}
We write $\Pbb(\Aset) = \int_{\Aset} f(x) dx$ for measurable $\Aset \subseteq \Rd$.
We 
let $\Csigr := \set{x: 0 < \dist(x, \Csig) < r}$, where $\Csig$ is as in Theorem \ref{thm: conductance_upper_bound}. 
Our goal will be to upper bound $p_K$ by a term that depends on the probability mass $\Pbb(\Csigr)$, 
and the bulk of our technical effort will be devoted to showing the following upper bound on $\Pbb(\Csigr)$:
%Lemma \ref{lem: expected_number_boundary_points} involves the bulk of the technical effort required to prove Theorem \ref{thm: conductance_upper_bound}; it will be necessary to bound the expected cut size of $\Csig[\Xbf]$ in $G_{n,r}$. 
\begin{lemma}
	\label{lem: expected_number_boundary_points}
	Under the conditions of Theorem \ref{thm: conductance_upper_bound}, and for any $0 < r \leq \sigma/2d$,
	\begin{equation*}
	\Pbb(\Csigr) \leq \frac{2dr}{\sigma} \left(\lambda_{\sigma} - c_0\frac{r^{\gamma}}{\gamma + 1}\right) \nu(\Csig)
	\end{equation*}	
\end{lemma}
\noindent Define the \emph{uniform local conductance} $\ell_{\nu,r}(u)$ to be
\begin{equation*}
\ell_{\nu,r}(u) = \nu\bigl(\Csig \cap B(u,r)\bigr).
\end{equation*}
In order to lower bound $p_V$ we will show the following lower bound on the uniform local conductance:
\begin{lemma}
	\label{lem: local_conductance}
	Let $u \in \Csig$. Then, for any $0 < r \leq \frac{\sigma}{2\sqrt{d}}$,
	\begin{equation*}
	\ell_{\nu,r}(u) \geq \frac{6}{25} \nu_d r^d.
	\end{equation*}
\end{lemma}
Taking these two results as given we can now prove each of the two claims~\eqref{eqn:claim_one} and~\eqref{eqn:claim_two} in turn.
\paragraph{Proof of Claim~\eqref{eqn:claim_one}: } Recalling,~\eqref{eqn: density_cut_expansion} 
	and by the linearity of expectations, we obtain
	\begin{equation*}
	p_K = \frac{\mu_K}{{n \choose 2}} = 2 \cdot \Pbb(x_i \not\in \Csig, x_j \in \Csig, \norm{x_i - x_j} \leq r). \tag{for each $i,j$, $i \neq j$}
	\end{equation*}
	Writing this with respect to the density function $f$, we have
	\begin{align*}
	p_K & = 2 \int_{\Rd \setminus \Csig} f(x) \Pbb\bigl(B(x,r) \cap \Csig\bigr) \dx \\
	& = 2 \int_{\Csigr} f(x) \Pbb\bigl(B(x,r) \cap \Csig\bigr) \dx \\
	& \leq 2 \nu_d r^d \lambda  \int_{\Csigr} f(x) \dx = 2 \nu_d r^d \lambda \Pbb(\Csigr).
	\end{align*}
	where the inequality follows from \ref{asmp: cluster_separation}, which implies $f(x) \leq \lambda$ for $x \in \Csig \setminus \Cset$. Then, upper bounding the integral using \sbcomment{I changed the Lemma referenced here. I think it was wrong before} Lemma \ref{lem: expected_number_boundary_points} gives the final result.

%\subsection{Proof of Lemma~\ref{Z}}
\paragraph{Proof of Claim~\eqref{eqn:claim_two}: } Recalling,~\eqref{eqn: volume_expansion}
%The proof will proceed similarly to Lemma \ref{lem: expected_density_cut}. 
	and by the linearity of expectation we obtain
	\begin{equation*}
	p_V = \frac{\mu_V}{{n \choose 2}} = 2 \cdot \Pbb(x_i \in \Csig, x_j \in B(x_i,r)). \tag{for any $i,j$, $i \neq j$. }
	\end{equation*}
	Writing this with respect to the density function $f$, we have
	\begin{align*}
	p_V & = 2 \int_{\Csig} f(x) \Pbb(B(x,r)) \dx \\
	& \geq 2 \int_{\Csig} f(x) \Pbb(B(x,r) \cap \Csig) \dx
	\end{align*}
	whence the claim then follows by Lemma \ref{lem: local_conductance}.
It remains to prove Lemmas~\ref{lem: expected_number_boundary_points} and~\ref{lem: local_conductance}, and we turn our attention to this next.
\subsection{Proof of Lemma~\ref{lem: expected_number_boundary_points}}
The proof of this Lemma relies on certain volume estimates whose statement and proof we defer to Appendix~\ref{sec: volume_estimates}.
	We partition $\Csigr$ into slices based on distance from $\Csig$ as follows: for $k \in \N$,
	\begin{equation*}
	\mathcal{T}_{i,k} = \set{x \in \Csigr: t_{i,k} < \frac{\dist(x, \Csig)}{r} \leq t_{i+1,k}}, ~~ \Csigr = \bigcup_{i = 0}^{k-1} \mathcal{T}_{i,k},
	\end{equation*}
	\sbcomment{Changed $t_i$ to $t_{i,k}$.}
	where $t_{i,k} = i/k$ for $i = 0, \ldots, k - 1$. As a result, for any $k \in \mathbb{N}$,
	\begin{equation}
	\label{eqn: partition_ub}
	\Pbb(\Csigr) = \int_{\Csigr} f(x) \dx = \sum_{i = 0}^{k-1} \int_{\mathcal{T}_{i,k}} f(x) \dx \leq \sum_{i = 0}^{k-1} \nu(\mathcal{T}_{i,k}) \max_{x \in \mathcal{T}_{i,k}} f(x).
	\end{equation}
	Assumptions~\ref{asmp: bounded_density} and \ref{asmp: low_noise_density} imply the upper bound
	\begin{equation*}
	\max_{x \in \mathcal{T}_{i,k}} f(x) \leq \lambda_{\sigma} - c_0(rt_{i,k})^{\gamma},
	\end{equation*}
	and writing
	\begin{equation*}
	\nu(\mathcal{T}_{i,k}) = \nu(\Csig + rt_{i+1,k}B) - \nu(\Csig + rt_{i,k}B) =: \nu_{i+1,k} - \nu_{i,k},
	\end{equation*}
	we have
	\begin{align}
	\label{eqn: telescoping_sum}
	\sum_{i = 0}^{k-1} \nu(\mathcal{T}_{i,k}) \max_{x \in \mathcal{T}_{i,k}} f(x) & \leq \sum_{i = 0}^{k-1} \biggl\{ \nu_{i+1,k} - \nu_{i,k} \biggr\} \biggl( \lambda_{\sigma} - c_0(rt_{i,k})^{\gamma} \biggr) \nonumber \\
	& = \underbrace{\sum_{i = 1}^{k} 
	\nu_{i,k} \biggl( \left[\lambda_{\sigma} - c_0(rt_{i-1,k})^{\gamma}\right] -  \left[\lambda_{\sigma} - c_0(rt_{i,k})^{\gamma}\right]\biggr)}_{:= \Sigma_k} + \underbrace{\biggl(\nu_{k,k}\left[\lambda_{\sigma} - c_0r^{\gamma}\right] - \nu_{1,k}\lambda_{\sigma} \biggr)}_{:= \xi}
	\end{align}
	\sbcomment{I changed the braces above -- I think they were incorrect earlier} where the second equality comes from rearranging terms in the sum.

@@

	and therefore \sbcomment{Added a subscript $\sigma$ on LHS}

@@

	\label{lem: covering}\sbcomment{Changed $A$ to $\Aset$ in the statement}

@@

\subsection{Concentration inequalities}
\label{sec: concentration}

Given a symmetric kernel function $K: \Rd \to \Reals$ with $\sup_{x,y \in \Rd} \abs{K(x,y)} \leq 1$, and independent and identically distributed data $\set{x_1, \ldots, x_n}$, we define the \textit{order-$2$ $U$-statistic} to be 
\begin{equation*}
U := \frac{1}{ {n \choose 2} } \sum_{1 \leq i < j \leq n} K(x_i,x_j).
\end{equation*}

@@

We will require large-sample and asymptotic bounds on several different degree and volume functionals, which follow from Lemmas \ref{lem: bounded_difference} and \ref{lem: multiplicative_Hoeffding}. Let
\begin{alignat*}{2}
\degminwt & := \min_{u \in \Csig[\Xbf]} \deg(u; G_{n,r}\left[\Csig[\Xbf]\right]),~ && \degmaxwt := \max_{u \in \Csig[\Xbf]} \deg(u; G_{n,r}\left[\Csig[\Xbf]\right]) \\ 
\degminpr & := \min_{u \in \Csig'[\Xbf]} \deg(u; G_{n,r}),~ && \degmax := \max_{u \in \Csig[\Xbf]} \deg(u; G_{n,r}) \\
\degmin & := \min_{u \in \Xbf} \deg(u; G_{n,r}).
\end{alignat*}

\begin{lemma}
	\label{lem: ball_bounds_in_probability}
	Under the geometric conditions \ref{asmp: bounded_density}-\ref{asmp: bounded_volume}, for any $\epsilon > 0$, each of the following bounds hold with probability tending to one as $n \to \infty$:
	\begin{align}
	\frac{\degminwt}{n} & \geq (1 - \epsilon)\frac{6}{25} \lambda_{\sigma} r^d \nu_d \label{eqn: min_cluster_degree_bound} \\
	\frac{\degmax}{n} & \leq (1 + \epsilon)\Lambda_{\sigma} r^d \nu_d \label{eqn: max_degree_bound}\\
	\frac{\abs{\Csig[\Xbf]}}{n} & \leq (1 + \epsilon)\Pbb(\Csig) \label{eqn: Csig_cardinality_bound}\\
	\label{eqn: volwt_bound}
	\frac{1}{2}\vol_0 &\leq \vol_{n,r}(\Csig[\Xbf])  \leq \frac{3}{2} \vol_0
	\end{align}

	Additionally, with probability one,
	\begin{align}
	\liminf_{n \to \infty} \frac{\degminwt}{n} & \geq \frac{6}{25} \lambda_{\sigma} r^d \nu_d \label{eqn: min_cluster_degree_bound_asymp}\\
	\limsup_{n \to \infty} \frac{\widetilde{\vol}_{n,r}(\Csig[\Xbf])}{n(n-1)} & \leq \Pbb(\Csig) \Lambda_{\sigma} \nu_d r^d \label{eqn: max_vol_bound_asymp}
	\end{align}

	If additionally \ref{asmp: valid_region} holds: for any $\epsilon > 0$, the following bound holds with probability tending to one as $n \to \infty$:
	\begin{align}
	\frac{\degmin}{n} \geq (1 - \epsilon)\frac{6}{25}\lambda_{\min} r^d \nu_d \label{eqn: min_degree_bound} 
	\end{align}

>>>>

\begin{theorem}
  \label{thm: consistent_recovery_of_density_clusters}
  Fix $\lambda > 0$, let $\Cset \in \Cbb_f(\lambda)$ be a
  $\kappa$-well-conditioned density cluster (with respect to some $\theta$), and
  additionally assume $f$ satisfies \ref{asmp: C'_bounded_density}. If Algorithm
  \ref{alg: ppr} is well-initialized, there exists a universal constant $c >  0$
  such that if  
  \begin{equation}
    \label{eqn: kappa_ub}
    \kappa(\Cset) \leq c \frac{\lambda_{\sigma}^2r^d
      \nu_d}{\Lambda_{\sigma}\Pbb(\Csig)}, 
  \end{equation}
  then the output set $\Cest \subseteq \Xbf$ is a consistent estimator for
  $\Cset$, in the sense of Definition \ref{def: consistent_density_cluster_estimation}. 
\end{theorem}
\noindent Some remarks are in order: 
 \begin{enumerate}  
 \item We note that the restriction on $\kappa(\Cset)$ imposed by \eqref{eqn:
    kappa_ub} results in a symmetric set difference \smash{$\Delta(\Csig[\Xbf],
    \Cest)$} on the order of $r^d$. In plain terms, we are able to recover a
  density cluster $\Cset$ in the sense of \eqref{eqn:
    consistent_density_cluster_recovery} only when we can guarantee a very small
  fraction of points will be misclassified. This strong condition is the price
  we pay in order to obtain the uniform bound in~\eqref{eqn: ppr_gap}. 
 \item While taking the radius of the neighborhood graph $r \to 0$ as $n \to
  \infty$ (thereby ensuring that the graph $G_{n,r}$ is sparse) is computationally
  attractive, the presence of a factor of \smash{$\log^2(1/r)/r$} in
  $\kappa(\Cset)$ prevents us from making claims about the
  behavior of PPR in this regime. Although the restriction to a kernel
  function fixed in $n$ is standard for spectral clustering theory
  \citep{schiebinger2015,vonluxburg2008}, it is an interesting question whether
  PPR exhibits some degeneracy over $r$-neighborhood graphs as $r \to 0$, or if
  this is merely looseness in our upper bounds.  
\end{enumerate}

@@

\subsection{Upper Bounds on the Mixing Time} 
For $S \subseteq V$, denote by $G[S] = (S, E_S)$ the
subgraph induced by $S$ (where the edges are $E_S = E \cap (S \times S)$). Let
$\Wbf_S$ be the (lazy) random walk matrix over $G[S]$, and write  
$$
q_{v}^{(t)}(u) = e_v\Wbf_S^t e_u
$$
for the $t$-step transition probability of the lazy random walk over $G[S]$
originating at $v \in V$. Also write \smash{$\pi = (\pi(u))_{u \in S}$}
for the stationary distribution of this random walk.  (As
$\Wbf_S$ is the transition matrix of a lazy random walk, 
it is well-known that a unique stationary distribution exists and is given by 
\smash{$\pi(u) = (\Dbf_S)_{uu}/\vol(S; G[S])$}, where we write $\Dbf_S$ for the  
degree matrix of $G[S]$.) We define the \emph{mixing time} of $G[S]$ as
\begin{equation}
\label{eqn: mixing_time}
\tau_{\infty}(G[S]) = \min\set{ t: \frac{\pi(u) - q_{v}^{(t)}(u)}
  {\pi(u)} \leq \frac{1}{4}, \; \text{for $u,v \in V$}}. 
\end{equation}
Next, we give an asymptotic (in the number of samples $n$) upper bound on
$\tau_{\infty}(G_{n,r}[\Csig[\Xbf]])$.

\begin{theorem}
  \label{thm: mixing_time_upper_bound}
  Fix $\lambda > 0$, and assume that \smash{$\Cset \in \Cbb_f(\lambda)$}
  satisfies Assumptions \ref{asmp: bounded_density} and \ref{asmp: embedding}
  for some $\sigma, \lambda_{\sigma}, \Lambda_{\sigma}, \rho, L > 0$. Then, for
  any \smash{$0 < r < \sigma/2\sqrt{d}$}, with probability 1,
  \begin{equation} 
    \label{eqn: mixing_time_upper_bound}
    \limsup_{n \to \infty}\tau_{\infty}(G_{n,r}[\Csig[\Xbf]]) \leq c_2 \frac{\Lambda_{\sigma}^4 d^3 \rho^2 L^2}{\lambda_{\sigma}^4 r^2} \log^2\left(\frac{c_4 \Lambda_{\sigma}}{\lambda_{\sigma}^2 r}\right) + c_3
  \end{equation}
  for $c_2,c_3,c_4 > 0$ universal constants. 
\end{theorem}

@@

\paragraph{Empirical behavior of PPR.} In Figure \ref{fig:fig2}, to drive home
the main implications of Theorems \ref{thm: misclassification_rate} and
\ref{thm: consistent_recovery_of_density_clusters}, we show the
behavior of PPR, normalized cut, and the density clustering algorithm of
\citet{chaudhuri2010} on the well known ``two moons'' dataset (with added 2d 
Gaussian noise), considered a prototypical success story for spectral clustering
algorithms. The first column shows the empirical density clusters $\Cset[\Xbf]$
and $\Cset'[\Xbf]$ for a particular threshold $\lambda$ of the density function; the 
second column shows the cluster recovered by PPR; the third column shows the
global minimum normalized cut, computed according to the algorithm of
\citet{szlam2010}; and the last column shows a cut of the density cluster tree
estimator of \citet{chaudhuri2010}.  We can see the degrading ability of PPR to
recover density clusters as the two moons become less well-separated. Of 
particular interest is the fact that PPR fails to recover one of the moons even
when normalized cut still succeeds in doing so. Additionally, we note that the Chaudhuri-Dasgupta algorithm succeeds even when both PPR and normalized cut fail.  While our main message was
that PPR recovers geometrically well-conditioned density clusters, it would be  
interesting to establish that it \emph{only} recovers such clusters, a direction
for future work.

@@

Taking these two technical lemmas as given we can now complete the proof of the theorem. 
We fix $\delta \in (0,1]$ and let $\delta' = \delta/3$. We rewrite
\begin{equation}
\label{eqn: conductance_representation_1}
\Phi_{n,r}(\Csig[\Xbf]) = \frac{p_K + \left(\frac{\cut_{n,r}}{{n \choose 2}} - p_K\right)}{p_V + \left(\frac{\min\set{\vol_{n,r}, \vol_{n,r}^c}}{{n \choose 2}} - p_V\right)}.
\end{equation}
By a union bound we see that all three bounds in \eqref{eqn: numerator_additive_bound} hold with probability at least $1 - \delta$. Along with (\ref{eqn: conductance_representation_1}) this yields the upper bound
\begin{equation*}
\Phi_{n,r}(\Csig[\Xbf]) \leq \frac{p_K + \Err_n}{p_V - \Err_n}
\end{equation*}
for $\Err_n = \sqrt{\frac{\log(1/\delta')}{n}}$.
Now, some straightforward algebraic manipulations yield,
\begin{align*}
\frac{p_K + \Err_n}{p_V - \Err_n} & = \frac{p_K}{p_V} \left(\frac{p_V}{p_V - \Err_n}\right) + \frac{\Err_n}{p_V - \Err_n} \\
& = \frac{p_K}{p_V} + \left(\frac{p_K}{p_V} + 1\right)\frac{\Err_n}{p_V - \Err_n} \\
& \leq \frac{p_K}{p_V} + 2 \frac{\Err_n}{p_V - \Err_n}.
\end{align*}

@@

\subsection{Proof of Lemma~\ref{lem: expected_density_cut}}
We write $\Pbb(\Aset) = \int_{\Aset} f(x) dx$ for measurable $\Aset \subseteq \Rd$.
We 
let $\Csigr := \set{x: 0 < \dist(x, \Csig) < r}$, where $\Csig$ is as in Theorem \ref{thm: conductance_upper_bound}. 
Our goal will be to upper bound $p_K$ by a term that depends on the probability mass $\Pbb(\Csigr)$, 
and the bulk of our technical effort will be devoted to showing the following upper bound on $\Pbb(\Csigr)$:
%Lemma \ref{lem: expected_number_boundary_points} involves the bulk of the technical effort required to prove Theorem \ref{thm: conductance_upper_bound}; it will be necessary to bound the expected cut size of $\Csig[\Xbf]$ in $G_{n,r}$. 
\begin{lemma}
	\label{lem: expected_number_boundary_points}
	Under the conditions of Theorem \ref{thm: conductance_upper_bound}, and for any $0 < r \leq \sigma/2d$,
	\begin{equation*}
	\Pbb(\Csigr) \leq \frac{2dr}{\sigma} \left(\lambda_{\sigma} - c_0\frac{r^{\gamma}}{\gamma + 1}\right) \nu(\Csig)
	\end{equation*}	
\end{lemma}
\noindent Define the \emph{uniform local conductance} $\ell_{\nu,r}(u)$ to be
\begin{equation*}
\ell_{\nu,r}(u) = \nu\bigl(\Csig \cap B(u,r)\bigr).
\end{equation*}
In order to lower bound $p_V$ we will show the following lower bound on the uniform local conductance:
\begin{lemma}
	\label{lem: local_conductance}
	Let $u \in \Csig$. Then, for any $0 < r \leq \frac{\sigma}{2\sqrt{d}}$,
	\begin{equation*}
	\ell_{\nu,r}(u) \geq \frac{6}{25} \nu_d r^d.
	\end{equation*}
\end{lemma}
Taking these two results as given we can now prove each of the two claims~\eqref{eqn:claim_one} and~\eqref{eqn:claim_two} in turn.
\paragraph{Proof of Claim~\eqref{eqn:claim_one}: } Recalling,~\eqref{eqn: density_cut_expansion} 
	and by the linearity of expectations, we obtain
	\begin{equation*}
	p_K = \frac{\mu_K}{{n \choose 2}} = 2 \cdot \Pbb(x_i \not\in \Csig, x_j \in \Csig, \norm{x_i - x_j} \leq r). \tag{for each $i,j$, $i \neq j$}
	\end{equation*}
	Writing this with respect to the density function $f$, we have
	\begin{align*}
	p_K & = 2 \int_{\Rd \setminus \Csig} f(x) \Pbb\bigl(B(x,r) \cap \Csig\bigr) \dx \\
	& = 2 \int_{\Csigr} f(x) \Pbb\bigl(B(x,r) \cap \Csig\bigr) \dx \\
	& \leq 2 \nu_d r^d \lambda  \int_{\Csigr} f(x) \dx = 2 \nu_d r^d \lambda \Pbb(\Csigr).
	\end{align*}
	where the inequality follows from \ref{asmp: cluster_separation}, which implies $f(x) \leq \lambda$ for $x \in \Csig \setminus \Cset$. Then, upper bounding the integral using Lemma \ref{lem: expected_number_boundary_points} gives the final result.

%\subsection{Proof of Lemma~\ref{Z}}
\paragraph{Proof of Claim~\eqref{eqn:claim_two}: } Recalling,~\eqref{eqn: volume_expansion}
%The proof will proceed similarly to Lemma \ref{lem: expected_density_cut}. 
	and by the linearity of expectation we obtain
	\begin{equation*}
	p_V = \frac{\mu_V}{{n \choose 2}} = 2 \cdot \Pbb(x_i \in \Csig, x_j \in B(x_i,r)). \tag{for any $i,j$, $i \neq j$. }
	\end{equation*}
	Writing this with respect to the density function $f$, we have
	\begin{align*}
	p_V & = 2 \int_{\Csig} f(x) \Pbb(B(x,r)) \dx \\
	& \geq 2 \int_{\Csig} f(x) \Pbb(B(x,r) \cap \Csig) \dx
	\end{align*}
	whence the claim then follows by Lemma \ref{lem: local_conductance}.
It remains to prove Lemmas~\ref{lem: expected_number_boundary_points} and~\ref{lem: local_conductance}, and we turn our attention to this next.
\subsection{Proof of Lemma~\ref{lem: expected_number_boundary_points}}
The proof of this Lemma relies on certain volume estimates whose statement and proof we defer to Appendix~\ref{sec: volume_estimates}.
	We partition $\Csigr$ into slices based on distance from $\Csig$ as follows: for $k \in \N$,
	\begin{equation*}
	\mathcal{T}_{i,k} = \set{x \in \Csigr: t_{i,k} < \frac{\dist(x, \Csig)}{r} \leq t_{i+1,k}}, ~~ \Csigr = \bigcup_{i = 0}^{k-1} \mathcal{T}_{i,k},
	\end{equation*}
	where $t_{i,k} = i/k$ for $i = 0, \ldots, k - 1$. As a result, for any $k \in \mathbb{N}$,
	\begin{equation}
	\label{eqn: partition_ub}
	\Pbb(\Csigr) = \int_{\Csigr} f(x) \dx = \sum_{i = 0}^{k-1} \int_{\mathcal{T}_{i,k}} f(x) \dx \leq \sum_{i = 0}^{k-1} \nu(\mathcal{T}_{i,k}) \max_{x \in \mathcal{T}_{i,k}} f(x).
	\end{equation}
	Assumptions~\ref{asmp: bounded_density} and \ref{asmp: low_noise_density} imply the upper bound
	\begin{equation*}
	\max_{x \in \mathcal{T}_{i,k}} f(x) \leq \lambda_{\sigma} - c_0(rt_{i,k})^{\gamma},
	\end{equation*}
	and writing
	\begin{equation*}
	\nu(\mathcal{T}_{i,k}) = \nu(\Csig + rt_{i+1,k}B) - \nu(\Csig + rt_{i,k}B) =: \nu_{i+1,k} - \nu_{i,k},
	\end{equation*}
	we have
	\begin{align}
	\label{eqn: telescoping_sum}
	\sum_{i = 0}^{k-1} \nu(\mathcal{T}_{i,k}) \max_{x \in \mathcal{T}_{i,k}} f(x) & \leq \sum_{i = 0}^{k-1} \biggl\{ \nu_{i+1,k} - \nu_{i,k} \biggr\} \biggl( \lambda_{\sigma} - c_0(rt_{i,k})^{\gamma} \biggr) \nonumber \\
	& = \underbrace{\sum_{i = 1}^{k} 
	\nu_{i,k} \biggl( \left[\lambda_{\sigma} - c_0(rt_{i-1,k})^{\gamma}\right] -  \left[\lambda_{\sigma} - c_0(rt_{i,k})^{\gamma}\right]\biggr)}_{:= \Sigma_k} + \underbrace{\biggl(\nu_{k,k}\left[\lambda_{\sigma} - c_0r^{\gamma}\right] - \nu_{1,k}\lambda_{\sigma} \biggr)}_{:= \xi}
	\end{align}
	where the second equality comes from rearranging terms in the sum.

@@

	and therefore 

@@

	\label{lem: covering}

@@

\section{Probabilistic bounds}
\label{sec: concentration}

Given a symmetric kernel function $K: \Rd \times \Rd \to \Reals$ with $\sup_{x,y \in \Rd} \abs{K(x,y)} \leq 1$, and independent and identically distributed data $\set{x_1, \ldots, x_n}$, we define the \textit{order-$2$ $U$-statistic} to be 
\begin{equation*}
U := \frac{1}{ {n \choose 2} } \sum_{1 \leq i < j \leq n} K(x_i,x_j).
\end{equation*}

@@

We will require large-sample and asymptotic bounds on several different degree and volume functionals, which follow from Lemmas \ref{lem: bounded_difference} and \ref{lem: multiplicative_Hoeffding}. Let $\Cset' \in \Cbb_f(\lambda)$ be a high-density cluster, with $\sigma$-expansion $\Csig' = \Cset' + \sigma B$, and recall the notation $\widetilde{\deg}_{n,r}(x) = \deg_{n,r}(x,G_{n,r}[\Csig[\Xbf]])$ for $x \in \Csig[\Xbf]$. Then, let
\begin{alignat*}{2}
\degminwt & := \min_{u \in \Csig[\Xbf]} \deg(u; G_{n,r}\left[\Csig[\Xbf]\right]),~ && \degmaxwt := \max_{u \in \Csig[\Xbf]} \deg(u; G_{n,r}\left[\Csig[\Xbf]\right]) \\ 
\degminpr & := \min_{u \in \Csig'[\Xbf]} \deg(u; G_{n,r}),~ && \degmax := \max_{u \in \Csig[\Xbf]} \deg(u; G_{n,r}) \\
\degmin & := \min_{u \in \Xbf} \deg(u; G_{n,r}).
\end{alignat*}
We collect the bounds we need in Lemma \ref{lem: ball_bounds_in_probability}.
\begin{lemma}
	\label{lem: ball_bounds_in_probability}
	Under the geometric conditions \ref{asmp: bounded_density}-\ref{asmp: bounded_volume}, for any $\epsilon > 0$, each of the following bounds hold with probability tending to one as $n \to \infty$:
	\begin{align}
	\frac{\degminwt}{n - 1} & \geq (1 - \epsilon)\frac{6}{25} \lambda_{\sigma} r^d \nu_d \label{eqn: min_cluster_degree_bound} \\
	\frac{\degmax}{n} & \leq (1 + \epsilon)\Lambda_{\sigma} r^d \nu_d \label{eqn: max_degree_bound}\\
	\frac{\abs{\Csig[\Xbf]}}{n} & \leq (1 + \epsilon)\Pbb(\Csig) \label{eqn: Csig_cardinality_bound}\\
	\label{eqn: volwt_bound}
	\frac{1}{2}\vol_0 &\leq \vol_{n,r}(\Csig[\Xbf])  \leq \frac{3}{2} \vol_0
	\end{align}
